---
title: "R coding Demo-- Fintech and Lending"
author: "Professor Lina Han"
output:
  html_document:
    df_print: paged
---

In this question, we will build a credit model and use it to predict consumer default loan defaults. Specifically,we will study how the performance of default prediction will change when we introduce more information available.

We will start with a sample database of 10,000 three-year consumer loans - "pslcdata.csv". The database contains the following details for each borrower: ID, Loan Amount, Interest Rate (annual in percentage), Credit Grade, Detailed Credit Grade, Length of Employment, Revolving Credit Balance, Revolving Credit Utility (in percentage), FICO score, Home Ownership Status, Annual Income, Employment Verification, Loan Status and Default status of the loan.

## 1: Basic Model Estimate - the case with FICO score only

In this part, we will work with *fico score only* and see how that predicts default.

(a) Transform variable default into a format that R can understand. (1 for default and 0 for non-default).

```{r}

# You can install and read in packages here
library(data.table)

# load data, change to your own data folder

setwd("C:/Users/linahan/OneDrive - University of Massachusetts/Teaching UMass/Teaching Fall 2025/R demo/Demo1_credit")
X <- read.csv("pslcdata.csv")

# lets have a look at the data
colnames(X)

# the default variable
unique(X$default)

# create a binary outcome variable
X$default_dummy[!X$default=="Defaulted"] <- 0
X$default_dummy[X$default=="Defaulted"] <- 1


# lets have a look at the data again
summary(X)

```

(b) Before estimating the model, what sign do we expect to see for the coefficient of fico score in a logistic regression? 


(c) Write an R script to estimate a *logistic regression* model with FICO score as predictor. 

```{r}
logit <- glm(data=X, formula=default_dummy~fico, family=binomial(link="logit"))
summary(logit)
```

## 2: Model Evaluation

Now that we have the first fitted model, we want to evaluate its usefulness.

(a) Use the model to estimate the *probability* of each loan's default.

```{r}

X$pred <- predict(object=logit,data=X, type="response")

head(X$pred)

```

(b) Use the probability estimated, create an *ROC* curve plot. Use the code provided below for solving this part of the problem.

```{r}
# there are many ways to do this a lot shorter, this is meant to be step-by-step

#install.packages("ROCR")
library(ROCR)

pred <- prediction(X$pred, X$default_dummy)
perf <- performance(pred,"tpr","fpr")

plot(perf,col="red")

abline(c(0,0),c(1,1))

```

(c) What is the area under the curve? Is the area below your *ROC* curve greater than 0.5? 

```{r}

auc_ROCR <- performance(pred, measure = "auc")
auc_ROCR@y.values[[1]]

```

(d) Suppose we develop a policy that rejects all the loan applicants whose predicted probability of default is greater than 10%. 
What is the proportion of consumers you correctly reject? (TPR,The denominator should be the total number of people who defaulted.)
What is the proportion of consumers you mistakenly reject? (FPR,The denominator should be total number of people who do not default.) 


```{r}
# create a variable with the default predictions 
X$pred_def <- F
X$pred_def[X$pred>=0.1] <- T

# number of people who default
print(paste("Number of defaults",sum(X$default_dummy)))

# true positives, i.e. people who were rejected and also defaulted
print(paste("True positives:",sum(X$default_dummy==1 & X$pred_def>=0.1)))

# true positive rate
print(paste("True positives:", sum(X$default_dummy==1 & X$pred_def>=0.1)/sum(X$default_dummy)))

# total number of people who do not default
num_neg <- sum(!X$default_dummy)
print(paste("Total number of people who do not default:",num_neg))

# false positives, i.e. people who where rejected at the 10% threshold but did not default
num_false_pos <- sum(!X$default_dummy[X$pred_def])
print(paste("False positive number:",num_false_pos))
print(paste("Proportion of people mistakenly rejected:",num_false_pos/num_neg))

```

## 3: With Slightly Richer Data

In this part, you are expected to conduct some further analysis taking advantage of the additional variables available. You are expected to run a logistic regression with *fico, loan amount, interest rate, detailed credit grade, revolving credit balance, employment verification status and Lending club's sub grade.* The employment verification should be treated as two groups: non-verified and everything else.

(a) Which variables should be transformed to factors in R?


```{r}

summary(X)
# converting category variables to dummy: verification status
X$verification <- 1L
X$verification[X$verification_status=="Not Verified"] <- 0L

```

(b) Run a logistic regression using the explanatory variables listed above. Report your logistic regression results.

```{r}
logit2 <- glm(data=X, formula=default_dummy~fico+loan_amnt+int_rate+revol_bal+verification, family=binomial(link="logit"))
summary(logit2)
```

(c) Create new default predictions from this richer model and plot an ROC curve.

```{r}
# new default probability estimates
X$pred1 <- predict(object=logit2, type="response")

# ROC curve
pred1 <- prediction(X$pred1, X$default_dummy)
perf1 <- performance(pred1,"tpr","fpr")

plot(perf,col = "red")
par(new=TRUE)
plot(perf1,col = "blue")

abline(c(0,0),c(1,1))




```

(d) Is the area below your new ROC curve greater than that in part 2? Report the area. Interpret the differences.

```{r}

auc_ROCR <- performance(pred1, measure = "auc")
auc_ROCR@y.values[[1]]

#print(auc)
```


